### 2018-04-05 latex/egsr17-cgf/conclusion.tex
updated paper

In this paper, we introduced the $P_N$-method to the toolbox of deterministic methods for rendering participating media in computer graphics. We derived and presented the real-valued $P_N$-equations, along with a staggered grid solver for solving them. Unique to our system is, that it uses a computer algebra representation of the equation to solve as input. Discretization in angular and spatial domain is done without any user interference, using manipulation passes on the representation. The stencil code, which is used to propagate the system of linear equations, is automatically generated from the expression tree.

This way, we can easily deal with complex coupled PDE's and avoid the time consuming and error prone process of writing stencil code by hand. Our approach helps especially for higher order, when the number of equations increase and become more complex.

We further show, how to use the results in a rendering system and run our solver for various standard problems for validation.

\begin{itemize}
	\item Mention performance problems and improving of solver performance as future work.
	\item mention vaccum convergence problem and the challenge to couple deterministic methods with surface rendering problems.
	\item Apply our automated stencil generation to other problem domains where coupled PDE's are present.
\end{itemize}

>>>>

In this paper, we introduced the $P_N$-method to the toolbox of deterministic methods for rendering participating media in computer graphics. We derived and presented the real-valued $P_N$-equations, along with a staggered grid solver for solving them. We showed, how to use the results in a rendering system and ran our solver for various standard problems for validation.

We originally set out to understand how non-linear diffusion compares against $P_N$ for increasing order. Although the lack of higher moments makes the FLD solution very flat, its energy conserving nature and comparably small resource footprint make it a much better approach when compared to $P_N$, which becomes more and more intractable for higher moment count. This is especially true, when compromises in accuracy of the approximation are acceptable.

The literature in other fields often states, that the $P_N$ method---when solved in normal form like we do---is able to deal with vacuum regions. This is misleading. The $P_N$-method in normal form indeed does not break down completely in the presence of vacuum like diffusion based methods do (due to $\sigma_t^{-1}$ in the diffusion coefficient). However, in the presence of vacuum, the condition number of the system matrix becomes infinite and does not converge either. Therefore $P_N$ based methods also require minimum thresholding of the extinction coefficient and offer no benefit for vacuum regions.

Much more work needs to be done in order to make our method competitive to alternative solutions. We believe this can be made possible by employing a multigrid scheme for solving the linear system of equations. We implemented a multigrid solver, but did not find the expected performance improvements. This is possibly due to the coupling between coefficients within a voxel, which does not work well together with the smoothing step. We want to study this further in the future.

Unique to our system is, that it uses a computer algebra representation of the equation to solve as input. Discretization in angular and spatial domain is done using manipulation passes on the representation. The stencil code, which is used to propagate the system of linear equations, is automatically generated from the expression tree. This way, we can easily deal with complex coupled PDE's and avoid the time consuming and error prone process of writing stencil code by hand.

When researching the application of $P_N$ in other fields, we came across a rich variety of variations on the $P_N$-method, such as simplified $P_N$ ($SP_N$)~\cite{Ryan10}, filtered $P_N$ ($FP_N$)~\cite{Radice13}, diffusion-correction $P_N$ ($DP_N$)~\cite{Schaefer11} and least-squares $P_N$ ($LSP_N$)~\cite{Hansen14}. These variations have been introduced to address certain problems of the standard $P_N$-method, such as ringing artefacts, dealing with vacuum regions and general convergence. Our solver can be applied to any (potentially coupled) PDE and therefore can generate stencil code for all these variations by simply expressing the respective PDE's in our computer algebra representation and providing this as an input to our solver. This would allow an exhaustive comparison of all these methods and we consider this as future work.

Finally, since the approach of our solver is very generic, we also would like to explore its application to other simulation problems in computer graphics.

### 2018-04-05 latex/egsr17-cgf/introduction.tex
updated paper

Deterministic methods on the other hand are unpopular. This is because they suffer from discretization artifacts, produce biased results, cannot be coupled easily with surface rendering problems and are trickier to implement. However, their appeal lies in the fact that they have better performance for certain problems and produce a global solution across the whole domain.

@@

The $P_N$-method is a deterministic method, which is popular in other fields like medical imaging and nuclear sciences, but has not been explored for applications in graphics yet. The purpose and main contribution of our paper is, to gain a solid understanding of its foundations and present a method for application in rendering. In particular, we present the following theoretical and practical contributions:
\begin{itemize}
	\item We derive and present the time-independent real-valued $P_N$-equations.
	\item We introduce a staggered-grid solver, for which we generate stencil code automatically from a computer algebra representation of the $P_N$-equations. This allows us to deal with the increasingly complex equations, which the $P_N$-method produces for higher order. It further allows our solver to be used for any (potentially coupled) partial differential equation, which results in a system of linear equations after discretization.
	\item Finally, we compare the $P_N$-method for higher orders against flux-limited diffusion, and answer the long standing question about which approach should be preferred.
\end{itemize}

In the next section, we will discuss related work and why it motivates our contribution. Section~\ref{sec:pnequations} introduces the real-valued $P_N$-equations, which we solve using our solver described in section~\ref{sec:pnsolver}. We explain in section~\ref{sec:rendering}, how the solution is being applied in a rendering application. Our $P_N$-solver is run and compared against flux-limited diffusion for a set of standard problems in section~\ref{sec:results}. We finish with our conclusion in section~\ref{sec:conclusion}.

>>>>

Deterministic methods on the other hand are unpopular, because they suffer from discretization artifacts, produce biased results, cannot be coupled easily with surface rendering problems and are trickier to implement. However, their appeal lies in the fact that they have better performance for certain problems and produce a global solution across the whole domain.

@@

The $P_N$-method is a deterministic method, which is popular in other fields like medical imaging and nuclear sciences, but has not been explored for applications in graphics yet. The purpose and main contribution of our paper is, to gain a solid understanding of its foundations and present a method for using it on problems in rendering. In particular, we present the following theoretical and practical contributions:
\begin{itemize}
	\item We derive and present the time-independent real-valued $P_N$-equations. To our knowledge, we are the first to derive and write down their compact form.
	\item We introduce a staggered-grid solver, for which we generate stencil code automatically from a computer algebra representation of the $P_N$-equations. This allows us to deal with the increasingly complex equations, which the $P_N$-method produces for higher order. It further allows our solver to be used for any (potentially coupled) partial differential equation, which results in a system of linear equations after discretization.
	\item Finally, we compare the $P_N$-method for higher orders against flux-limited diffusion and ground truth Monte Carlo integration.
\end{itemize}

In the next section, we will discuss related work and why it motivates our contribution. In section~\ref{sec:discretized_rte} we revisit the deterministic approach to light transport simulation in participating media and outline the discretization using spherical harmonics. In section~\ref{sec:car} we introduce our computer algebra representation, which we required to derive the real-valued $P_N$-equations. This representation is also a key component of our solver, which we present in section~\ref{sec:pnsolver}. We explain, in section~\ref{sec:rendering}, how the solution is being applied in a rendering application. Our $P_N$-solver is run and compared against flux-limited diffusion for a set of standard problems in section~\ref{sec:results}. We finish with our conclusion in section~\ref{sec:conclusion}.

### 2018-04-05 latex/egsr17-cgf/main-body.tex
updated paper

\title[Draft]%
{
%Evaluation of Higher Order Diffusion against Variable Eddington Factor Methods for Rendering Participating Media
A Deterministic Method for Simulation of Light Transport in Participating Media based on Higher Order Spherical Harmonics
}

@@

\begin{abstract}
%Classical diffusion approximation (CDA) is very competitive for rendering highly scattering participating media, because in these scenarios, Monte Carlo based methods suffer from high computational cost. Variable Eddington Factor methods, such as flux-limited diffusion (FLD), have shown to further improve the results from CDA. In this paper, we look at CDA as a degenerate case of the $P_1$-method and investigate the open question, if going higher order gives a viable alternative to FLD. We derive the time-independent real-valued $P_N$-equations and introduce a framework for generating finite difference stencils directly from the computer algebra representation of the $P_N$-equations for arbitrary order. We further present a staggered-grid $P_N$-solver and compare it against FLD for various standard problems.
Rendering highly scattering participating media is a common challenge for brute-force path tracing. Classical diffusion approximation is a competitive alternative, as it reduces the problem to solving a simple diffusion equation. Extensions, such as flux-limited diffusion, improve the former by switching to a non-linear diffusion coefficient. Both methods are based on the spherical harmonics expansion of the radiance field, truncated after the first order. In this paper, we investigate the open question of wether higher orders provide a viable alternative to these two approaches. Increasing the order introduces a set of increasingly complex coupled partial differential equations, whose growing number and complexity make them very difficult to work with for higher order. We introduce a computer algebra framework for representing and manipulating the underlying mathematical equations and use it to derive the time-independent real-valued $P_N$-equations for arbitrary orders. We further present a staggered-grid $P_N$-solver and generate its stencil code directly from the expression tree of the $P_N$-equations. Finally, we discuss how our method compares against prior work for various standard problems. We will release our computer algebra system, solver, and data as open source to ensure full reproducibility of all of our results.
%\begin{classification} % according to http://www.acm.org/class/1998/
%\CCScat{Computer Graphics}{I.3.7}{Three-Dimensional Graphics and Realism}{Raytracing}
%\end{classification}
\end{abstract}

@@

%-------------------------------------------------------------------------
\input{pn_equations.tex}

>>>>

\title[Draft]%
{
%Evaluation of Higher Order Diffusion against Variable Eddington Factor Methods for Rendering Participating Media
$P_N$-Method for Multiple Scattering in Participating Media
}

@@

\begin{abstract}
%Classical diffusion approximation (CDA) is very competitive for rendering highly scattering participating media, because in these scenarios, Monte Carlo based methods suffer from high computational cost. Variable Eddington Factor methods, such as flux-limited diffusion (FLD), have shown to further improve the results from CDA. In this paper, we look at CDA as a degenerate case of the $P_1$-method and investigate the open question, if going higher order gives a viable alternative to FLD. We derive the time-independent real-valued $P_N$-equations and introduce a framework for generating finite difference stencils directly from the computer algebra representation of the $P_N$-equations for arbitrary order. We further present a staggered-grid $P_N$-solver and compare it against FLD for various standard problems.
Rendering highly scattering participating media is a common challenge for brute-force path tracing. Classical diffusion approximation is a competitive alternative, as it reduces the problem to solving a simple diffusion equation. Extensions, such as flux-limited diffusion, improve the former by switching to a non-linear diffusion coefficient. Both methods are based on the spherical harmonics expansion of the radiance field, truncated after the first order. In this paper, we investigate the open question of wether higher orders provide a viable alternative to these two approaches. Increasing the order introduces a set of increasingly complex coupled partial differential equations, whose growing number and complexity make them very difficult to work with. We introduce a computer algebra framework for representing and manipulating the underlying mathematical equations and use it to derive the time-independent real-valued $P_N$-equations for arbitrary orders. We further present a staggered-grid $P_N$-solver and generate its stencil code directly from the expression tree of the $P_N$-equations. Finally, we discuss how our method compares against prior work for various standard problems. We will release our computer algebra system, solver, and data as open source to ensure full reproducibility of all of our results.
%\begin{classification} % according to http://www.acm.org/class/1998/
%\CCScat{Computer Graphics}{I.3.7}{Three-Dimensional Graphics and Realism}{Raytracing}
%\end{classification}
\end{abstract}

@@

%-------------------------------------------------------------------------
\input{section_discretized_rte.tex}
\input{section_car.tex}
\input{section_real_valued_pn_equations.tex}

### 2018-04-05 latex/egsr17-cgf/previous_work.tex
updated paper

@@

The idea behind VEF is to replace the diffusion coefficient in the diffusion equation of CDA by a non-linear diffusion coefficient. This new diffusion coefficient is being derived by seeing it as a mean to spatially blend between the diffusive and pure transport regime. The diffusive regime is where thick highly scattering media is present and causes photons to change directions in short succession, while the transport regime in thin and highly absorbing media causes photons to travel long straight lines. In the absence of scattering (e.g. pure vacuum), the non-linear diffusion coefficient turns the diffusion equation into an advection equation.

It is an open and unresolved question, whether the $P_N$-method with truncation at higher order, gives any benefit over using first order non-linear diffusion. This question is also been raised in other domains~\cite{Olson00} and resolving it is the main motivation for this paper.

>>>>

\begin{figure}[h]
\centering
\begin{subfigure}{0.45\columnwidth}
%\missingfigure{test}
\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_collocated.png}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
%\missingfigure{test2}
\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_staggered.png}
\end{subfigure}%
\vspace{-0.1in}
\icaption{Solving the 2D-checkerboard problem using naive collocated grids produces oscillating artefacts (left). Our solver uses staggered grids and produces artefact free results (right).}
\label{fig:artifacts}
\end{figure}

@@

The idea behind VEF is to derive a better diffusion coefficient by seeing it as a mean to spatially blend between the diffusive and pure transport regime. The diffusive regime is where thick highly scattering media is present and causes photons to change directions in short succession, while the transport regime in thin and highly absorbing media causes photons to travel long straight lines. In the absence of scattering (e.g. pure vacuum), the non-linear diffusion coefficient turns the diffusion equation into an advection equation.

It is an open and unresolved question, whether the $P_N$-method with truncation at higher order, gives any benefit over using first order non-linear diffusion. It is also been raised in other domains~\cite{Olson00} and resolving it is one motivation for this paper.

### 2018-04-05 latex/egsr17-cgf/section_car.tex
updated paper

\section{$P_N$-Solver}
\label{sec:pnsolver}

Expanding, discretizing and working with the $P_N$-equations gets increasingly complex for higher order and is arduous and error prone if done by hand. We tackle this problem by expressing the RTE in a computer algebra representation and apply the discretization in angular and spatial domain using manipulation passes on the representation.

The discretized $P_N$-equations are then rendered into source code and compiled. We refer to this as the stencil code. It allows fast construction of the linear system $A\vec{u}=\vec{Q}$ for arbitrary resolution and RTE parameters and has no dependency on the computer algebra framework anymore.

Finally, the generated linear system is solved using standard methods. The solution $\vec{u}$ represents the discretized radiance field $\widehat{L}$, which is evaluated using trilinear interpolation of the SH coefficients. An overview of the solver is given in figure~\ref{fig:pnsolver}. Each step will be described in more detail in the following subsections.

\subsection{Computer Algebra Representation}

@@

\begin{figure}[h]
\centering
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test2}
\end{subfigure}%

\begin{subfigure}{0.45\columnwidth}
\missingfigure{test}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test2}
\end{subfigure}%
\vspace{-0.2in}
\icaption{The computer algebra representation is constructed using API commands provided by our framework (top left) and represented internally as an expression tree (top right). Additional API commands (bottem left) allow valid transformations, which modify the expression tree accordingly (bottom right).}
\end{figure}

Finally, frontends allow rendering the expression tree into different forms. We implemented frontends for rendering expression trees to \LaTeX and C++ source code. The equations in the supplemental material were almost all rendered by the former. The latter was used for generating the stencil code used by our solver.

@@

\subsection{Discretization}

The result of the discretization in angular domain are the real-valued $P_N$-equations. We arrived at the consice form in equations~\ref{eq:rpn_m_<_z}-~\ref{eq:rpn_m_>_z} using our framework. However, once the compact form was found, we used it directly as input to our solver.

The discretization code iterates over all $l,m$-pairs up to order $N$ and chooses either equation~\ref{eq:rpn_m_<_z}, equation~\ref{eq:rpn_m_=_z} or equation~\ref{eq:rpn_m_>_z}, according to the sign of $m$. All occurances of $l$ and $m$ are replaced with their actual values.

The spatial discretization is done by parsing the expression tree of each equation from the root. All occurances of the continuous positional variable $\vec{x}$ are replaced by discrete voxel coordinates $i,j,k$, whose values are retrieved from the top of a stack. The stack is kept by the parser and is initialized with the voxel at $(0,0,0)$. We refer to the space of the discretized $P_N$-equations as stencil space.

Whenever the parser encounters a differential operator, the current voxel position is shifted and pushed to the stack. Then the child expression is parsed and multiplied with a weight value, followed by popping the stack. This is done for each element of a central difference stencil, which provides the offset and weight values (referencing the voxelsize variable). The dimension of the stencil is given by the derivation variable of the operator. Nested derivatives are handled naturally and produce higher order finite difference stencils as expected.

Another aspect to consider during discretization is the location of the variables and unknowns. As we will see later, our solver needs to support staggered grids, where the SH coefficients are not all located at the voxel center (collocated). Some also can be located at the faces, edges and intersections of voxels (staggered).

Our solver therefore supports placement of coefficients at arbitrary staggered grid locations. Whenever a coefficient is encountered during expression tree parsing for discretization, the parser checks how its location is related to the current location on the top of the stack. Depending on how the two are located to each other, the parser returns an expression, which interpolates the coefficients at the requested offset from its surrounding locations, or it returns the coefficient itself, if it happens to coincide. This is also done for RTE parameters, such as $\sigma_t$ or $p^{l,m}$, which are always located at the voxel center. The position stack of the parser is initialized with the staggered grid location defined by the coefficient, which is associated with the $l,m$-pair of the equation currently being discretized.

\subsection{Stencil Code Generation and System Building}

After applying the discretization step to the expression tree of the $P_N$-equations, it is used to generate the stencil code. In numerical analysis, a stencil is an arrangement of voxels and weights, that relate values at different locations to each other and form the basis for propagating rows in the system matrix $A$ and RHS vector $\vec{Q}$ with values. The name comes from the fact, that the geometric structure and weights of the configuration do not change, when applied to different voxels. The same is true for the $P_N$-equations and we use this fact to generate a single function, which propagates rows in $A$ and $\vec{Q}$ for a given voxel.

The $P_N$-equations express, how each coefficient of a voxel depends on other coefficients of the same or adjacent voxels. The unknowns in the terms give information about the coefficient index and voxel offset, and therefore identify a column offset in the matrix $A$. The factors to these coefficients may contain evaluations of RTE parameters, such as $\sigma_t$. Therefore, these factors can not be determined during stencil generation, but are rendered into code expressions, which are executed as part of the stencil function during runtime. Because the stencil code has been generated in stencil space relative to the voxel at $(0,0,0)$, we can run the same stencil code for every voxel, by simply applying an offset accordingly.

The $P_N$-equations have been discretized, and the stencil has been generated, without any notion of domain boundaries and boundary conditions (BC). Our solver supports Neumann BC and Dirichlet BC. They are handled transparently by the code which runs the stencil. Whenever the stencil writes into $A$ for a boundary coefficient, the framework will either ignore the write operation (Dirichlet BC) or write into the row and column in $A$ of the same coefficient at the closest voxel inside the domain (Neumann BC).

In order to respect the boundary correctly, additional coefficients (which contribute additional rows and columns in $A$ and $\vec{Q}$) are necessary at boundary voxels (see Figure~\ref{fig:staggeredgrid}). This requires careful managegment and bookkeeping of coefficient indices, which is done transparently by the code running the stencil.

\begin{figure}[h]
\centering
\begin{subfigure}{0.45\columnwidth}
%\missingfigure{test}
\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_collocated.png}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
%\missingfigure{test2}
\includegraphics[width=\columnwidth]{images/checkerboard2d_p1_staggered.png}
\end{subfigure}%
\vspace{-0.1in}
\icaption{The solution of our solver for the 2D-checkerboard problem with collocated grids (left), where all coefficients are located at the cell centers. It suffers from severe oscillating artifacts, which are removed by offsetting the coefficient locations using staggered grids (right).}
\label{fig:artifacts}
\end{figure}

\subsection{Solving}

After running the stencil code for every voxel, the system matrix $A$
and RHS vector $\vec{Q}$ are propagated with values and can be used with standard methods for solving systems of linear equations. The number of rows is determined by the number of voxels times the number of coefficients per voxel and can therefore become very large for high resolution and high truncation order. The matrix $A$ is squared and fortunately very sparse, due to finite differences and the structure of the $P_N$-equations. Unfortunately, it also is non-symmetric (due to the transport term) and not diagonal dominant, which rules out many standard methods for solving linear systems. We adress this by solving the normal form $A^TA\vec{u}=A^T\vec{Q}$ instead. This gives a symmetric and positive definit system matrix $A^TA$, albeit with a higher condition number.

\begin{figure}[h]
\centering
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test}
\end{subfigure}%
\hspace{0.05\columnwidth}
\begin{subfigure}{0.45\columnwidth}
\missingfigure{test2}
\end{subfigure}%
\vspace{-0.2in}
\icaption{Staggered grids cause a shifted boundary on the right and upper side of the domain (left). The boundary is correct after introducing additional unknowns (red) at boundary voxels (right).}
\label{fig:staggeredgrid}
\end{figure}

%Matrix $A$ and vector $\vec{Q}$ are constructed after applying the stencil for every voxel.

%\subsection*{CDA vs. $P_1$}
%\begin{itemize}
  %\item CDA is a degenerated form of $P_1$. It is derived by isolating the flux-vector on one side of the vector-equation formed by the $l=1$ SH-band equations. This isolation requires division by the extinction coefficient, introducing a $\frac{1}{\sigma_t}$ factor which diverges as $\sigma_t$ approaches zero. Thresholding to some minimum is required, in order to be able to solve the system for vacuum regions.
  %\item $P_1$ does not require any thresholding of $\sigma_t$ as it does not contain $\sigma_t$ as a denominator. It therefore can deal with vacuum regions without modifications.
  %\item (needs validation) Further, in the presence of vacuum or near vacuum regions, the condition number for CDA is higher than for $P_1$, because of small extinction values in the denominator of the diffusion coefficient.
  %\item Using the normal form for CDA will further increase the condition number when vacuum regions are present and significantly decreases convergence.
%\end{itemize}

>>>>

\section{Computer Algebra Representation}
\label{sec:car}

@@

Finally, frontends allow rendering the expression tree into different forms. We implemented frontends for rendering expression trees to \LaTeX~and C++ source code. The equations in the supplemental material were almost all rendered by the former. The latter was used for generating the stencil code used by our solver.

@@
